\documentclass[a4paper,12pt]{article}

\usepackage{geometry}
\usepackage{cite}
\usepackage{url}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{hyperref}
\geometry{margin=1.5in}
\usepackage{enumitem}

\title{Efficient instruction-level parallelism analysis using tropical semiring \lq{matrix multiplication\rq}.}
\author{Bryan Abi Karam}
\date{\today}
\begin{document}

\maketitle

\section{Introduction}

% WHAT IS ILP?

Instruction-level parallelism (ILP) is a technique commonly
used by high-performance processors to improve performance.
Previous studies \cite{11096386} have investigated the theoretical
limits of ILP in programs using models of their dynamic execution.

% WHAT IS ONE UPPER BOUND FOR PARALLELISM? CRITICAL PATH

An upper bound for parallelism in a program arises from its critical
path: the longest chain of dependent instructions in the program's
dynamic execution. For a machine that takes at least one cycle per
instruction, the program cannot be run faster than the time it takes
to run the critical path sequentially.

Therefore an upper bound for ILP in a program can be calculated as the
number of dynamic instructions in the program's execution divided by the
length of the critical path.

% HOW TO COMPUTE LENGTH OF CRITICAL PATH?

The dependencies of instructions in a program execution can be represented
by its full dynamic dependency graph (DDG) \cite{753330}, which is a directed
acyclic graph with one node per dynamically executed instruction in a program.
The edges of the graph represent dependencies between dynamically executed instructions,
and so the length of the critical path is the length of the longest path through
the program's DDG.

% WHAT IS TROPICAL ANALYSIS?

Tropical analysis is a branch of pure mathematics using the tropical semiring \cite{Chadderz},
also known as the min/max-plus algebra, which is the semiring
\[(\mathbb{R} \cup \{+\infty\}, \oplus, \otimes)\]
with operations
\[
\begin{aligned}
  a \oplus b & = \min(a, b) \text{ or} \max(a, b) \\
  a \otimes b & = a + b
\end{aligned}
\]
The problem of finding the longest path in a DAG can be expressed as a series
of \lq{matrix multiplications\rq} in the max tropical semiring, as shown by
Section I-B of Chadwick \cite{Chadderz}.

% WHAT WILL I DO IN THIS PROJECT?

In this project I will create a library for performing such matrix
calculations efficiently, considering aspects such as vectorisation,
tiling, and representation sparsity. I will work with CPU based compute
and may work with GPU based compute as an extension.

\textit{This introduction is adapted from the abstract of Chadwick \cite{Chadderz}.}

\section{The project}

\subsection{Project core}
\label{sec:core}
First, a CPU-based library for performing individual matrix computations in the tropical
semiring should be implemented, using a dense representation of matrices.
Techniques such as vectorisation and tiling should be considered for performance gains.

Matrix multiplication expressions will be generated from a program execution's dependency
structure as a graph, with one node per register per dynamic instruction,
in the column-based representation proposed in Section I-B of Chadwick \cite{Chadderz}.

The aforementioned matrix computation library will be used to compute the results
of these expressions, using techniques such as caching and evaluation order optimisation
to maximise efficiency.

Finally, a benchmark suite for testing performance of implementations will be implemented,
and a baseline performance will be established using a \lq{classical\rq} technique for computing
dependency chains (i.e. simulating one instruction at a time naively, computing its dependencies
one-by-one).

\subsection{Project extensions}
\label{sec:extensions}

\subsubsection{Sparse matrix representation}
The core of the project implements a matrix computation library using a dense
representation of matrices. In an effort to improve memory usage, and potentially
improve performance (as the cost of sparse multiplication is proportional to the
number of non-zeroes in the matrix, rather than just the number of rows and columns),
I will implement a similar CPU-based matrix computation library using a sparse
representation of matrices as an extension to the project core.

The sparse library will be benchmarked against the dense library for register-only
program execution dependency graphs.

\subsubsection{GPU compute}
The core of the project implements a CPU-based matrix computation library.
To exploit the excellent parallel power of GPUs, I will implement two GPU-based matrix
computation libraries, for dense and sparse representations of matrices, using a
programming model such as OpenCL or CUDA.

The GPU libraries will be benchmarked against their CPU counterparts for register-only
program execution dependency graphs.

\subsubsection{Through-memory dependencies}
So far we have only computed results of matrix expressions generated from register-only
program execution dependency graphs. However, for real programs, the true critical path
often involves memory operations.

A simple approach is to naively include memory locations as graph nodes alongside
registers for each dynamic instruction, but this results in a catastrophic explosion of
matrix sizes. Nevertheless, these matrices are extremely sparse, as most memory operations
access only two or three locations at a time, and so are usable with sparse representations.
For this approach, I will benchmark the performance of the two previously implemented sparse
matrix computation libraries (running on CPU and GPU) against the baseline.

An alternative approach is to model the effects of memory operations using tropical semiring
\textit{addition} rather than just multiplication, as described in Section II-B of Chadwick \cite{Chadderz}.
For this approach, I will benchmark the performance of the two previously implemented dense
matrix computation libraries (running on CPU and GPU) against the baseline.

\section{Starting point}
Existing standard libraries for matrix multiplication cannot be used as they are
specialised for linear algebra (as opposed to tropical algebra).
Additionally, techniques such as Strassen's algorithm do not apply in a semiring.

A suite of tools will be provided by Alexandra Chadwick, the project supervisor,
to generate matrix expressions from program executions.
The tools will do the following:
\begin{itemize}
  \item Compute the static dependencies of every static instruction in a program.
  \item Turn the above into a matrix per instruction.
  \item Dump the dynamic flow of execution of a program.
  \item Use the dynamic flow and the per-instruction matrices to create a matrix expression.
\end{itemize}
Generated matrix expressions will be used as input to the matrix computation libraries.

\section{Success criteria and evaluation plan}

\subsection{Success criteria}

For the dissertation to be considered a success, all items in the project core
(\hyperref[sec:core]{Section 2.1}) should be completed, specifically the following items:
\begin{itemize}
  \item Implement a CPU-based library for performing individual matrix
    computations in the tropical semiring, using a dense representations of matrices.
  \item Use the CPU-based library to compute the result of large matrix expressions
    for computing longest paths in register-only DDGs.
  \item Establish a baseline performance using a classical technique for computing
    dependency chains.
  \item Implement a benchmark suite for testing performance of implementations.
\end{itemize}
For the core to be considered a success, the CPU-based library should demonstrate a
performance improvement over the naive baseline for computing the length of the critical path.

If time permits, I will complete the three project extensions described in
\hyperref[sec:extensions]{Section 2.2}, prioritising them in the order they are presented.
The success criteria for each extension is as follows:
\begin{itemize}
  \item Sparse matrix representation: The implementation from this extension should
    demonstrate a reduction in memory usage and an improvement in performance for very
    sparse matrices over the CPU-based dense implementation from the project core.
  \item GPU compute: The implementations from this extension should show
    performance improvement over their CPU-based counterparts.
  \item Through-memory dependencies: The benchmarks from this extension should
    demonstrate a performance improvement for the matrix computation libraries over
    the naive baseline for computing the length of the critical path in a full DDG.
\end{itemize}

All implementations of the matrix computation library should evaluate matrix
expressions \textbf{correctly}.

\subsection{Evaluation plan}

The evaluation of the matrix computation libraries will be conducted using a
comprehensive set of benchmarks on both register-only DDGs and ones with through-memory
dependencies (addition and sparse representations).
All implementations will be benchmarked against the naive baseline (sequential
dependency chain computation) to quantify performance improvements.

For each benchmark, the following metrics will be collected:
\begin{itemize}
  \item \textbf{Execution time}: Total time taken to compute the longest path, or
    evaluate the matrix expression.
  \item \textbf{Memory usage}: Peak and average memory consumption during computation.
  \item \textbf{Correctness}: Verification that computed results match expected
    results for known cases.
\end{itemize}

Results will be analysed to determine which approach is most effective for each
situation, and to validate the success criteria detailed above.

\section{Plan of work}

\subsection*{10/10-24/10: Implement naive baseline}
Doing this before working with matrix computations will allow me to understand
the starting state of dependency chain computation.

Deliverable: Naive baseline.

\subsection*{25/10-7/11: Implement CPU-based dense matrix computation library}
I will first implement the library for indiviual matrix multiplications,
focusing on performance optimisations.

Deliverable: Individual matrix computation library implemented.

\subsection*{8/11-21/11: Use CPU-based library to compute matrix expressions}
I will use the aforementioned library in computing the results of large matrix expressions.
Work from this sprint may spill over into the next sprint due to an assessment deadline for
the Category Theory module I am taking.

Deliverable: Library for computing matrix expressions implemented.

\subsection*{22/11-5/12: Catch up and benchmark suite}
First I will complete any outstanding work from the previous sprint, and develop a benchmark suite
for testing solution performance in preparation for evaluation.

Deliverable: Benchmark suite implemented.

\subsection*{6/12-19/12: Evaluation of core}
I will evaluate the results of the project core against the success criteria using the benchmark
suite developed in the previous sprint.

Deliverable: Evaluation of core completed.

\subsection*{20/12-2/1: Low-progress buffer}
As this sprint falls over the holiday season, the amount of work done may be minimal.
I will use this time as a buffer for any outstanding work to be completed.

\subsection*{3/1-16/1: Sparse matrix representation}
This is the first of the project extensions.

Deliverable: Matrix computation library using sparse representation implemented.

\subsection*{17/1-30/1: GPU-based matrix computation libraries}
This is the second of the project extensions.
Due to an assessment deadline for the Category Theory module I am taking over the first
week of Lent term, work from this sprint may spill over into the next sprint.

Deliverable: Matrix computation library using GPU compute implemented.

\subsection*{31/1-13/2: Catch up and progress report}
I will catch up on any outstanding work, and dedicate time to preparing the progress
report and presentation due on 06/02/2025.

\subsection*{14/2-27/2: Evaluation of first two extensions}
I will evaluate the results of the first two project extensions against the success criteria.

Deliverable: Evaluation of first two extensions completed.

\subsection*{28/2-13/3: Investigate through-memory dependencies}
The second half of this sprint will coincide with an assessment deadline for the Multicore
Semantics and Programming module I am taking, and so progress may be limited.

This is the third of the project extensions.
This extension is an investigation of the performance of the previously implemented
libraries on full DDGs, and so may not require as much work as the previous extensions.

Deliverable: Investigation of through-memory dependencies completed.

\subsection*{14/3-27/3: Catch up and evaluate last extension}
The first half of this sprint will coincide with an assessment deadline for the Multicore
Semantics and Programming module I am taking, and so progress may be limited.

This sprint will be used to catch up on any outstanding work, and complete evaluations for all
extensions.

Deliverable: Evaluation of all extensions completed.

\subsection*{28/3-10/4: Start write-up}
In this sprint I will plan the dissertation structure, and start writing
chapters 1 (introduction) and 2 (preparation) of the dissertation for review.

If time permits, I will start writing chapter 3 (implementation).

Deliverable: Dissertation document plan drafted, and chapters 1 and 2 written.

\subsection*{11/4-24/4: Continue write-up}
I will address feedback received on chapters already written, then I will
continue writing the dissertation, aiming to complete chapters 3 (implementation),
4 (evaluation), and 5 (conclusions).

Deliverable: Dissertation chapters 3, 4 and 5 written.

\subsection*{25/4-8/5: Submit dissertation}
I will address feedback on all the wriitng done in previous sprints, and then
proofread the dissertation to be ready for submission 1 week before the deadline of
15/05/2025.

\section{Resource declaration}
% List resources required for the project (hardware, software, data, etc.)
TBC

\bibliography{bibliography}{}
\bibliographystyle{siam}
\end{document}
