\chapter{Conclusions}

The project was overall a success. The analysis tool built offers a direct performance improvement over the classical 
techniques for conducting ILP limit studies. All the minimum success criteria were met, and nearly all the extensions have 
been implemented with varying levels of success. The associativity of the matrix approach has been shown to offer 
clear pathways for both data-driven and parallel optimisations.

\section{Lessons learned}

The theory behind the matrix-based approach was well-established and understood going into the project.
However, pattern exploitation was not well-defined in the project proposal, and so much 
of the work in Section \ref{impl:fast-expr} was exploratory and took more time 
than expected. This left little time to deal with the steep learning curves of multithreaded and GPU programming, and 
highlighted the importance of domain knowledge and tool preparation. Additionally, these hardware optimisations were 
developed and optimised locally on my laptop, and not iteratively tested on suitable hardware. For the multithreaded
approach, this meant that utilising more cores on the server hardware was not impactful enough to overtake the fewer, 
yet more powerful laptop cores.

Evaluation proved to be a tricky part of the project. While its importance was understood, my methodology was 
initially underdeveloped and so I was unable to thoroughly evaluate my progress as I made changes.
Instead, much of the project was benchmarked all at once during the writeup, and this made fixing critical performance
bugs more difficult.

\section{Future work}

Listed are shortcomings of the analysis tool that could be addressed in the future.

\begin{itemize}
  \item The GPU pipeline did not make full use of the available compute. The memory layout could be optimised for
    contiguous storage, a double-buffered dispatch architecture implemented to hide overhead, and 
    a hybrid system built that simultaneously multiplies matrices on CPU threads to maximise throughput.
  \item The tool's memory usage is unbounded and problematic.
    More work could be done to ensure that the maximum memory usage is safely limited 
    during analysis of realistic traces of trillions of dynamic instructions.
  \item I/O and parsing form a sequential bottleneck for the analysis tool when loading the initial expression. 
    An alternative binary-based format could be developed to speed up loading and reduce the bottleneck.
  \item The tool only tracked register dependencies. While this is still useful for smaller subtraces that keep their 
    working set in registers, programs will often spill registers to memory. This occurred with all the sample traces,
    and so reported ILP limits were higher than actual values. The impact of the data-driven optimisations could be
    assessed on more complex program traces involving memory dependencies.
\end{itemize}

Overall, the project achieved its goals, and I think it was an interesting project, bridging gaps between concepts in 
processor design, graph theory, compression and parallel computing.
