\chapter{Formatting Checklist}

\LaTeX documents are easy to format well with little effort. If you're coming from a `what you see is what you get' word processor, use the following (non-exhaustive) checklist to check your report is neatly formatted. Finding out how to use these formatting styles is mainly left as an exercise to the reader.

\begin{itemize}
    \item Are your speech marks correctly formatted?  
    \begin{itemize}
        \item Example: A ``common mistake'' is for users to incorrectly use normal quotation mark characters in their \LaTeX\ files.
        \item Anti-Pattern Example: Some think that computer science is "boring".
    \end{itemize}

    \item Are you using the \href{https://nhigham.com/2019/11/19/better-latex-tables-with-booktabs/}{\texttt{booktabs}} package for neater tables?  

    \item Are you citing correctly?  
    \begin{itemize}
        \item Example: Kitchen supplies are known to go missing in an office context \citep{CaseOfDisappearingTeaspoons}. However,  \citet{TeaspoonSolutionProposal} has proposed an elegant solution.
        \item Anti-Patterns:
        \begin{itemize}
            \item Not putting spaces between text and reference, e.g., Kitchen supplies are known to go missing in an office context\citep{CaseOfDisappearingTeaspoons}.
            \item Incorrect use of \texttt{\textbackslash citep\{\}} or \texttt{\textbackslash citet\{\}}, e.g., Kitchen supplies are known to go missing in an office context \citet{CaseOfDisappearingTeaspoons}. However, \citep{TeaspoonSolutionProposal} has proposed an elegant solution.
        \end{itemize}
    \end{itemize}

    \item Are references to programmatic things (e.g., file, class, or variable names) in a monospaced font?
    \begin{itemize}
        \item Use \texttt{\textbackslash texttt\{\}} for this
    \end{itemize}

    \item Are code extracts formatted as code listings (see Appendix \ref{app:latex-primer})?
    
    \item Is the text in your figures readable?  

    \item Are all of your paragraphs evenly spaced?  

    \item Are you using math mode when using equations/symbols?  

    \item Are your captions for figures appearing below figures and captions for tables appearing above them?
    \begin{itemize}
        \item Anti-Patterns:
        \begin{itemize}
            \item Putting captions in wrong place.
            \item Not including captions at all.
        \end{itemize}
    \end{itemize}

    \item Are your titles, section headings, and captions in sentence case?
    \begin{itemize}
        \item Example: \textbf{My new type system for Rust}
        \item Anti-pattern: \textbf{My New Type System For Rust}
    \end{itemize}

    \item Are your footnotes appearing immediately after a word?
    \begin{itemize}
        \item Example: This program was written in Python\footnote{Version 3.12 at the time of writing}.
        \item Anti-Pattern: This experiment was conducted in January 2025 \footnote{Apart from one participant, who took part in April 2025.}.
    \end{itemize}

    \item Are there sufficient details in your references?
    \begin{itemize}
        \item Example: See the detail of references included in \citet{CaseOfDisappearingTeaspoons}.
        \item Anti-Pattern: Missing key information such as the year published, all of the authors of the work, the DOI if applicable, etc. See \citet{BadReference} for an example of a poor reference.
    \end{itemize}

\end{itemize}